---
title: "Data Exploreation test"
author: "ningzhibin@gmail.com"
date: "`r Sys.time()`"
output:
  html_document:
    fig_width: 10
    fig_caption: TRUE
    toc: true
    toc_depth: 4
    toc_float:
      collapsed: TRUE
      smooth_scroll: TRUE
    number_sections: true
---


<style type="text/css">
.main-container {
  max-width: 1800px;
  margin-left: auto;
  margin-right: auto;
}
</style>


```{r include =FALSE}
# version 0.5
# 20190613
# fixed a lot of bugs
# a lot of layout and comments
```




```{r setup, include = FALSE}
# this doc could be used a as general service report template 

# enviroment setup
knitr::opts_chunk$set(echo = FALSE,warning = FALSE, message = FALSE, cache = TRUE)
# if you have a large file (large dataset to process) to knit, set cache = TRUE to speed up the test process

library(tidyverse)
library(DT) # for interactive data table display 


```



```{r data_preparation, eval=TRUE}

####################################################################
# parameter setting
####################################################################




# read in table, and match the gene name, then save the file
# read-in the expression file
df_pg <- read_tsv("Sample20-ProteinGroups(2)(1).txt", col_names  =  TRUE)


# for consistency,change the name
df_pg_exp_GSEA <- df_pg[,c(-1,-2,-3)]
df_pg_ID_exp_GSEA <- df_pg[,c(-2,-3)]

#setup the metadata
meta <- read.delim("Experiment_group.txt",header = FALSE)
colnames(meta) <- c("sample_Name", "Experiment_group")


# setup the two groups to compare
#groups_VS <- c("CK", "PCK")
#groups_VS <- c("CK", "HC")
groups_VS <- c("HC", "PHC")

#groups_VS <- c("PCK", "PHC")


```


===============================================================================================================

# Experimental Design


## data description


 **There are `r length(unique(meta[,1]))` samples: **
 
 `r levels(unique(meta[,1]))`

 **There are `r length(unique(meta[,2]))` groups: **
 
 `r levels(unique(meta[,2]))`


```{r table meta}
knitr::kable(meta, caption = 'Experimental design')
```


===============================================================================================================

## Expression Description

 **Overall raw intensity**


```{r plot display of raw intensity}
matrix_ggboxplot(as.matrix(df_pg_exp_GSEA), xlabel="Sample", ylabel = "Intensity", maintitle = "Intensity")$violinplot

#plot_distribution$boxplot
#plot_distribution$violinplot
```


===============================================================================================================

 **Overall log transformed intensity**

```{r plot display of log(intensity)}

plot_distribution <- matrix_ggboxplot(log10(as.matrix(df_pg_exp_GSEA)), xlabel="Sample", ylabel = "log10(Intensity)", maintitle = "Log10 Intensity")
plot_distribution$boxplot
plot_distribution$violinplot
```


===============================================================================================================

## heatmap and clustering 

  Display with the log10 intensity from all proteins


 **The following is from the data without scaling on rows(proteins)**

```{r heatmap without scaling}
matrix_quick_heatmap(log10(as.matrix(df_pg_exp_GSEA) +1 ),scale = "none", Col_tree = FALSE)

```


 **The following is from the data with scaling on rows(proteins)**


```{r}
matrix_quick_heatmap(log10(as.matrix(df_pg_exp_GSEA)),scale = "row", Col_tree = FALSE)

```


# PCA 

 **For overall data examination**




```{r PCA}
data_matrix <- log10(as.matrix(df_pg_exp_GSEA))
#rownames(data_matrix) <- 1:nrow(data_matrix) # it has to have row.names

pca_plsda <- PCA_PLSDA_wrapper(data_matrix = data_matrix+0.01, data_meta =  meta)
pca_plsda$plot_PCA_Scree
pca_plsda$plot_PCA_component


```

**Interactive 3D plot**

```{r}
pca <-PCA_wrapper_prcomp2(data_matrix, data_meta =  meta, inputation = TRUE)
pca$pca_component_plot_3d_interactive
```


# PLSDA 



 **Analysis to for key variable selection**

### Overall performance

```{r}
pca_plsda$plot_PLSDA_component
pca_plsda$plot_PLSDA_AUC

```

### heatmap with VIP filtering

PLSDA provides a powerful way to select the "important" features/variables to seperate the groups.
But do PLSDA is a supervised classification, with a great tendency of over fitting. 
DO NOT use the PLSDA result as PCA result. The grouping/classification does not mean the real differnce between samples. 



**heamap without VIP filtering**

```{r fig.width= 15,fig.height=10}
pca_plsda$plot_PLSDA_heatmap

```

**heamap with VIP >1**

```{r fig.width= 15,fig.height=10}
pca_plsda$plot_PLSDA_VIP_heatmap
```

### VIP filtering

 **VIP**


Variable importance in projection (VIP) coefficients reflects the relative importance of each X variable for each Y variate in the prediction model. VIP coefficients thus represent the importance of each X variable in fitting both the X- and Y-variates, since the Y-variates are predicted from the X-variates.

VIP allows to classify the X-variables according to their explanatory power of Y. Predictors with large VIP, larger than 1, are the most relevant for explaining Y.


```{r}
pca_plsda$plot_PLSDA_VIP
```



 **Table shows All VIPs for the first 3 componets**
 


```{r}
datatable(pca_plsda$data_PLSDA_VIP)
```


 
 **Table shows `r nrow(pca_plsda$data_PLSDA_VIP_filtered_features)` the features with VIP score (of the 1st component) > 1**
 
 These proteins/features can also be deemed as significantly changed proteins/features, and be used for downstream function analysis


```{r display_PLSDA_VIP}

#datatable(pca_plsda$data_PLSDA_VIP_filtered_features)
datatable(df_pg[as.numeric(row.names(pca_plsda$data_PLSDA_VIP_filtered_features)),])



```








# ANOVA

p value threshold is set to 0.05


```{r ANOVA}

# take out the expression matrix, change the name to the major protein id, for downstream statistics
df_pg_exp <- df_pg_exp_GSEA

pvalues_anova <- matrix_PostHoc(df_pg_exp,meta$Experiment_group)

df_pg_exp_p <- cbind(pvalues_anova, df_pg_exp_GSEA)

df_pg_exp_p_filtered <- df_pg_exp_p[which(!is.na(df_pg_exp_p$p_PostHoc_pairs)),] 


```

## Result

The following table shows the proteins with anova pvalue < 0.05 

 **Table shows `r dim(df_pg_exp_p_filtered)[1]` features which has/have at least one pvalue between any two experimental condidtions**


The first column shows the pvalue from the corresponding two groups.


```{r}
datatable(as.matrix(df_pg_exp_p_filtered))
#write.table(df_pg_exp_p, "proteinGroups_ttest.txt", sep = "\t",  quote = FALSE,col.names = NA, row.names =  TRUE)
```


# tTest between `r groups_VS[1]` and `r groups_VS[2]`


**The ttest and fold change are caclulated from the raw intensity, not the log10 value**

```{r}

matrix_fold_change<-function(data,group_info,group_name1,group_name2) {
  group1<-which(group_info==group_name1)
  group2<-which(group_info==group_name2)

  fold_change<-apply(data,1,function(x) sum(x[group1])/sum(x[group2]))
  
  return(fold_change)
}

# calculate the statistics
fold_change <- matrix_fold_change(df_pg_exp,meta$Experiment_group,groups_VS[1],groups_VS[2])

p <- matrix_ttest(df_pg_exp,meta$Experiment_group,groups_VS[1],groups_VS[2])
p_adjusted <- p.adjust(p,method = "fdr")


# combine all p value and selected columns together
df_pg_exp_ttest <- cbind(fold_change,p,p_adjusted,df_pg_exp_GSEA[,c(which(meta$Experiment_group==groups_VS[1]), which(meta$Experiment_group==groups_VS[2]))])

row.names(df_pg_exp_ttest) <- as.data.frame(df_pg_ID_exp_GSEA)[,1]


# rank the table based on the p value
df_pg_exp_ttest <- df_pg_exp_ttest[order(df_pg_exp_ttest$p_adjusted),]

df_pg_exp_ttest_p_05 <- df_pg_exp_ttest[df_pg_exp_ttest$p < 0.05,]


```



**There are `r length(which(p <0.05))` protein/features with p < 0.05 between `r groups_VS[1]` and `r groups_VS[2]`**


**There are `r length(which(p_adjusted <0.05))` with p_adjusted < 0.05 after FDR correction.**



**The table shows proteins with p<0.05, not p_adjust, which is too stringent for this dataset**

```{r}



datatable(df_pg_exp_ttest_p_05)

#df_pg_exp_p_CK_PCK <- cbind(fold_change_CK_PCK,p_CK_PCK_adjusted,df_pg_exp_GSEA)
#significant_CK_PCK <- df_pg_exp_p_CK_PCK[p_CK_PCK_adjusted < 0.05 ,]
#datatable(df_pg_exp_p_CK_PCK)
```



==========================================================================================================


